Numerical Investigation of Viscous Flow Instabilities in Multiphase Heterogeneous Porous Media

### 2018-05-13 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC_Feb.tex
Adding files that hopefully were used in the original submission back on Feb2018

\medskip
Finite element basis functions for velocity and pressure fields are introduced in the discretisation of force-balance equations. Hybrid basis functions are also used to allow CV-based velocity to be extrapolated across the entire element. The extended Darcy equation (Eqn.~\ref{eqn:darcy_eqn}) is discretrised using a FE representation of $\mathbf{v}_{\alpha}=\underline{\underline{\sigma}}_{\alpha}\mathbf{u}_{\alpha}$ and $p$ with FE basis functions $Q_{j}$ and $P_{j}$, respectively -- note that $\underline{\underline{\sigma}}_{\alpha}$ lies in both CV and FEM spaces. Each component of the weak form of the extended Darcy equation is tested with the $\mathbf{v}_\alpha$ basis function, $Q_{j}$, to obtain:
\begin{eqnarray}
  \sum\limits_{E} \left. \int\limits_{\Omega_E} { {Q}}_{i} \left({\mathbf v}_\alpha + \nabla p  -{\mathbf s}_{u_\alpha} \right) dV \right. + \displaystyle \oint_{\Gamma_{E}} {Q}_i {\mathbf n} \left(p - \tilde{p}\right) d\Gamma + \nonumber \\ 
  \oint_{\Gamma_{\Omega}} {Q}_i {\mathbf n} \left(p - p_\text{bc}\right) d\Gamma = \bm{0},
                   \label{force-semi-disc} 
\end{eqnarray} 
where $\Omega_E$ and $\Gamma_{E}$ are the volume and boundary of element $E$, respectively, and $\Gamma_{\Omega}$ is the boundary of the computational domain. The numerical pressure $\tilde{p}$ appearing in the jump condition (second term in Eqn.~\ref{force-semi-disc}) is the arithmetic mean of the potentially discontinuous pressure across the element $E$ $\left(\text{consistent with the \PNDG[n]{m} element-pairs}\right)$. This term vanishes when a continuous formulation is used to discretise the pressure field $\left(\text{with \PN[n]{m} element-pairs}\right)$. The last term in Eqn.~\ref{force-semi-disc} is used to weakly enforce the pressure level to $p_\text{bc}$ on a computational domain boundary.
%Finite element basis functions for velocity and pressure fields are introduced in the discretisation of force-balance equation (Eqn.~\ref{eqn:darcy_eqn}). Hybrid basis functions are also used to allow CV-based velocity to be extrapolated across the entire element. $\underline{\underline{\sigma}}_{\alpha}$ is an absorption-like term that represents the implicit linearisation of the viscous frictional forces, and lies in both CV and FEM spaces.
%\begin{displaymath}
%   \underline{\underline{\sigma}}_{\alpha} = \displaystyle\frac{\mu_{\alpha}S_{\alpha}}{{\mathbf K}\mathcal{K}_{r\alpha}}.
%\end{displaymath}

@@

\medskip
From Eqns.~\ref{eq:MR}-\ref{Eqn:CoreyBrooks2}, it is clear that during fluid displacement, the mobility ratio changes as phase saturation, $S_{w}\left(x_{i},t\right)$ and $S_{nw}\left(x_{i},t\right)$, varies in time and space (Eqn.~\ref{eqn:saturation_eqn}). Therefore, with no lack of generality, the MR can be reduced to the viscosity ratio (VR),
\begin{displaymath}
    \text{VR} = \frc{\mu_{i}}{\mu_{j}},
\end{displaymath}
 that will be used in the parametrisation of the numerical simulations conducted in Section \ref{section:results}. Analysis performed in the following sections will make use of this simplified definition as phase saturation $\left(S_{\alpha}\right)$ is a time- and spatial-dependent prognostic field which is calculated along with pressure ($p$) and velocity $\left(\mathbf{u}_{\alpha}\right)$ fields.

@@

\medskip
Fluid flow dynamics through heterogeneous porous media are sensibly more complex than in homogeneous media and strongly depend on problem properties. Studies by \citet{langtangen_1992} on fluid flow dynamics in heterogeneous porous media (analytical and numerical solutions) demonstrated that hyperbolic Buckley-Leverett model is inherently unstable, \ie the expected uniform interface front collapses as any perturbation in physical parameters are imposed to the problem. Spatial variation in geological formations occurs in all length-scales, where heterogeneity characteristics in small length-scales (\ie pore) are statistically embedded into permeability (absolute and relative) and porosity parameters. Such multi-scale heterogeneity induces preferential flow pathways and plays a significant role in the onset of fluid instabilities as it triggers fingers formation and their accelerated growth \citep[see][]{ewing_1989,tchelepi_1994}.

@@

%\medskip
%As outlined in Section \ref{section:ViscousInstabilities}, viscous flow instabilities are characterised by spontaneous formation and growth of fluid elongations, called viscous fingers. Geometrical evolution in time and space of these elongations provides good indication of instabilities at the fluid front interface. 
\medskip
The computational domain with permeability distribution of 10$^{-4}\le\mathbf{K}\le$10$^{5}$  milidarcy (Fig.~\ref{fig:3DChannel_PermMesh})  was designed to naturally capture preferential flow pathways through channels of large absolute permeability. Pure (wetting) fluid is added into the domain with uniform velocity of 2.5 cm.s$^{-1}$ from the left-hand side face. No-flux boundary conditions are applied to all borders, except at the inlet (left-hand side) and outflow (right-hand side) faces. Initially, the domain was partially saturated with non-wetting fluid, \ie $S_{nw}(t=0)=0.80$, which was continuously displaced by the wetting fluid as the simulation progress. The domain was discretised with 235k tetrahedral \PNDG[1]{1} element-pairs.

\medskip
As demonstrated in the preliminary model validation (Section~\ref{section:results_initial_model_validation}), crossflow towards the upper region of the initial part of the domain is due to boundary pressures and diffusivities (both at different vertical layers), \ie caused by large permeability gradients (Fig.~\ref{fig:3DChannel_sat}). As the simulation progresses, non-wetting fluid is driven from regions with lower fluid resistance (central channel), although adjacent region (with low permeability) remains stagnant.

\medskip
Such preferential flow pathways are clearly indicated in Fig.~\ref{fig:3DChannel_planes}, where Darcy velocity vectors overlapped with wetting phase fluid saturation are shown. Fluid displacement occurs mainly in regions of relatively larger permeability, leaving other parts of the domain (\ie regions with low permeability distribution) with little or no momentum, indicating that at these regions the non-wetting fluid (\ie fluid 2) is not effectively displaced.

\medskip
In Sections \ref{section:results_homo_hete}-\ref{section:results_hete_fix_adapt}, finger's formation, growth and branching were triggered by domain heterogeneity represented by permeability differences. The progress of the saturation front and formation and growth of fingers in time and space can be seen in more details in isosurfaces (Fig.~\ref{fig:3DChannel_Isosurf}).

@@

\medskip
However, due to low-order accuracy of element-pairs -- \PNDG[1]{1} instead of \PN[1]{2} of previous sections \citep[for full investigation on numerical accuracy associated with these element-pairs, see][]{salinas2015,salinas_2016,salinas_2018,adam_2016, gomes_2017}, and coarser mesh throughout the computational domain, simulation performed with fixed mesh showed larger smeared fluid saturation.

@@

\medskip
Simulations involving formation and growth of viscous fingers were conducted with the CVFEM-based multi-fluid flow simulator Fluidity/IC-FERST model. The model is based upon a dual consistent pressure-velocity representation in CV and FEM spaces and a novel family of FE-pairs, \PN[n]{m} and \PNDG[n]{m}. Saturation and other scalar fields are discretised in CV space and may be fully discontinuous across finite element boundaries. In order to strongly enforce discretised Darcy equations at the boundaries between elements and control volumes, a directional-weighted flux-limited was introduced to take into account discontinuities risen from both, numerical formulation (based on discontinuous-Galerkin finite element method, DGFEM) and control volumes within finite elements.

\medskip
%(2)Summary (1-3 paragraphs) of the main results with the appropriate analysis (\ie how you work has proved the theory);
The numerical formulation was initially validated against laboratory experiments \citep{dawe_2008} to qualitatively assess model functionality to simulate crossflow (\ie preferential flow pathways) in a chequerboard permeability domain. Formation and development of viscous fingers in Hele-Shaw cells (homogeneous domain) were investigated to demonstrate that the onset of interface flow instability occur when VR$\ge$3. Maximum dendritic finger widths obtained from numerical simulations performed with VR=10 and VR=150 largely agree with expected parametric analytic solutions \citep{mclean_1981,guan_2003}. However, when perturbations are imposed into the domain's geophysical properties, \ie spatial variability of absolute permeability, numerical solutions partially agree with analytic solutions (Table~\ref{table:Heleshaw}). This may be due to the domain homogeneity assumption in the analytic solution (Eqns.~\ref {eqn:guaneqn}).

\medskip
Determining the early-onset flow instability is critical to accurately simulate formation, growth and coalescence of fingers, and therefore to: (a) predict displacement (sweep) efficiency and (b) design strategies to either mitigate or avoid fingering and preferential flow pathways. Mesh resolution is thus crucial to investigate Saffmann-Taylor instabilities, and adaptive dynamic mesh technology has proved to be an efficient tool to capture initial interface perturbations. Numerical simulations were performed with VR=10 using fixed and adaptive \PN[1]{2} element-pairs. Results (Figs.~\ref{fig:5regions_plottimenodeselements} and~\ref{fig:5regions_AdaptZoom}) demonstrated that by dynamically adapting the mesh to follow saturation gradients (\ie fluids interfaces), early-onset flow instability and further formation of long multi-fluid fingers, which have themselves split into sub-fingers, can be readily captured (see Figs.~\ref{fig:5regions_fixedmesh}i and~\ref{fig:5regions_adaptmesh}i). Overall computational cost of the simulation conducted with dynamically adapting the mesh were significantly smaller than the simulation conducted with fixed mesh as indicated by the number of elements/nodes (Figs.~\ref{fig:5regions_plottimenodeselements}).

\medskip
Investigation of preferential flow pathways was the focus of Section~\ref{section:results_3D}, where 3D numerical simulations of multi-fluid flow through semi-pervious / impervious channelised domain were performed. Numerical data indicated preferential flow through regions of larger permeability leaving adjacent region (with low permeability) with nearly stagnant fluids (Figs.~\ref{fig:3DChannel_sat}-~\ref{fig:3DChannel_planes}). Order of accuracy of the element-pair and mesh resolution play key-roles to capture directional flow pathways and fingers in 3D simulations.

\medskip
% (3)(optional) 1 paragraph of potential future work. Make sure that (1) and (2) are covered in the present Section.
This work focuses on complexity associated with flow channelling and instability dynamics in heterogeneous porous media. Methods used to simulate such flows are based in a novel high-order CVFEM formulation that accurately preserves sharp fluid saturation gradients associated with contrasting permeability distribution. Future work will include numerical investigation of (a) crossover from capillary fingering (at low fluid velocity) to viscous fingering for immiscible unstable flows, and (b) coupled viscous and density fingering in miscible fluid flows.

@@

\clearpage 
%% References with bibTeX database:
\bibliographystyle{elsarticle-harv} 
\bibliography{references}

@@

\listoftables
\clearpage
\input{article_table}
\clearpage  
\listoffigures
\clearpage

\input{article_figure1}
%\input{article_figure2}

>>>>

Finite element basis functions for velocity and pressure fields are introduced in the discretisation of force-balance equations. Hybrid basis functions are also used to allow CV-based velocity to be extrapolated across the entire element. The extended Darcy equation (Eqn.~\ref{eqn:darcy_eqn}) is discretrised using a FE representation of $\mathbf{v}_{\alpha}=\underline{\underline{\sigma}}_{\alpha}\mathbf{u}_{\alpha}$ and $p$ with FE basis functions $Q_{j}$ and $P_{j}$, respectively -- note that $\underline{\underline{\sigma}}_{\alpha}$ lies in both CV and FEM spaces. Each component of the weak form of the extended Darcy equation is tested with the $\mathbf{v}_\alpha$ basis function, $Q_{j}$, to obtain:
\begin{eqnarray}
  \sum\limits_{E} \left. \int\limits_{\Omega_E} { {Q}}_{i} \left({\mathbf v}_\alpha + \nabla p  -{\mathbf s}_{u_\alpha} \right) dV \right. + \displaystyle \oint_{\Gamma_{E}} {Q}_i {\mathbf n} \left(p - \tilde{p}\right) d\Gamma + \nonumber \\ 
  \oint_{\Gamma_{\Omega}} {Q}_i {\mathbf n} \left(p - p_\text{bc}\right) d\Gamma = \bm{0},
                   \label{force-semi-disc} 
\end{eqnarray} 
where $\Omega_E$ and $\Gamma_{E}$ are the volume and boundary of element $E$, respectively, and $\Gamma_{\Omega}$ is the boundary of the computational domain. The numerical pressure $\tilde{p}$ appearing in the jump condition (second term in Eqn.~\ref{force-semi-disc}) is the arithmetic mean of the potentially discontinuous pressure across the element $E$ $\left(\text{consistent with the \PNDG[n]{m} element-pairs}\right)$. This term vanishes when a continuous formulation is used to discretise the pressure field $\left(\text{with \PN[n]{m} element-pairs}\right)$. The last term in Eqn.~\ref{force-semi-disc} is used to weakly enforce the pressure level to $p_\text{bc}$ on a computational domain boundary.
%Finite element basis functions for velocity and pressure fields are introduced in the discretisation of force-balance equation (Eqn.~\ref{eqn:darcy_eqn}). Hybrid basis functions are also used to allow CV-based velocity to be extrapolated across the entire element. $\underline{\underline{\sigma}}_{\alpha}$ is an absorption-like term that represents the implicit linearisation of the viscous frictional forces, and lies in both CV and FEM spaces.
%\begin{displaymath}
%   \underline{\underline{\sigma}}_{\alpha} = \displaystyle\frac{\mu_{\alpha}S_{\alpha}}{{\mathbf K}\mathcal{K}_{r\alpha}}.
%\end{displaymath}

@@

From Eqns.~\ref{eq:MR}-\ref{Eqn:CoreyBrooks2}, it is clear that during fluid displacement, the mobility ratio changes as phase saturation, $S_{w}\left(x_{i},t\right)$ and $S_{nw}\left(x_{i},t\right)$, varies in time and space (Eqn.~\ref{eqn:saturation_eqn}). Therefore, with no lack of generality, the MR can be reduced to the viscosity ratio (VR),
\begin{displaymath}
    \text{VR} = \frc{\mu_{i}}{\mu_{j}},
\end{displaymath}
 that will be used in the parametrisation of the numerical simulations conducted in Section \ref{section:results}. Analysis performed in the following sections will make use of this simplified definition as phase saturation $\left(S_{\alpha}\right)$ is a time- and spatial-dependent prognostic field which is calculated along with pressure ($p$) and velocity $\left(\mathbf{u}_{\alpha}\right)$ fields.

@@

\medskip
Fluid flow dynamics through heterogeneous porous media are sensibly more complex than in homogeneous media and strongly depend on problem properties. Studies by \citet{langtangen_1992} on fluid flow dynamics in heterogeneous porous media (analytical and numerical solutions) demonstrated that hyperbolic Buckley-Leverett model is inherently unstable, \ie the expected uniform interface front collapses as any perturbation in physical parameters are imposed to the problem.

Spatial variation in geological formations occurs in all length-scales, where heterogeneity characteristics in small length-scales (\ie pore) are statistically embedded into permeability (absolute and relative) and porosity parameters. Such multi-scale heterogeneity induces preferential flow pathways and plays a significant role in the onset of fluid instabilities as it triggers fingers formation and their accelerated growth \citep[see][]{ewing_1989,tchelepi_1994}.

@@

%\medskip
%As outlined in Section \ref{section:ViscousInstabilities}, viscous flow instabilities are characterised by spontaneous formation and growth of fluid elongations, called viscous fingers. Geometrical evolution in time and space of these elongations provides good indication of instabilities at the fluid front interface.

The computational domain with permeability distribution of 10$^{-4}\le\mathbf{K}\le$10$^{5}$  milidarcy (Fig.~\ref{fig:3DChannel_PermMesh})  was designed to naturally capture preferential flow pathways through channels of large absolute permeability. Pure (wetting) fluid is added into the domain with uniform velocity of 2.5 cm.s$^{-1}$ from the left-hand side face. No-flux boundary conditions are applied to all borders, except at the inlet (left-hand side) and outflow (right-hand side) faces. Initially, the domain was partially saturated with non-wetting fluid, \ie $S_{nw}(t=0)=0.80$, which was continuously displaced by the wetting fluid as the simulation progress. The domain was discretised with 235k tetrahedral \PNDG[1]{1} element-pairs.

As demonstrated in the preliminary model validation (Section~\ref{section:results_initial_model_validation}), crossflow towards the upper region of the initial part of the domain is due to boundary pressures and diffusivities (both at different vertical layers), \ie caused by large permeability gradients (Fig.~\ref{fig:3DChannel_sat}). As the simulation progresses, non-wetting fluid is driven from regions with lower fluid resistance (central channel), although adjacent region (with low permeability) remains stagnant.

Such preferential flow pathways are clearly indicated in Fig.~\ref{fig:3DChannel_planes}, where Darcy velocity vectors overlapped with wetting phase fluid saturation are shown. Fluid displacement occurs mainly in regions of relatively larger permeability, leaving other parts of the domain (\ie regions with low permeability distribution) with little or no momentum, indicating that at these regions the non-wetting fluid (\ie fluid 2) is not effectively displaced.

In Sections \ref{section:results_homo_hete}-\ref{section:results_hete_fix_adapt}, finger's formation, growth and branching were triggered by domain heterogeneity represented by permeability differences. The progress of the saturation front and formation and growth of fingers in time and space can be seen in more details in isosurfaces (Fig.~\ref{fig:3DChannel_Isosurf}).

@@

However, due to low-order accuracy of element-pairs -- \PNDG[1]{1} instead of \PN[1]{2} of previous sections \citep[for full investigation on numerical accuracy associated with these element-pairs, see][]{salinas2015,salinas_2016,salinas_2018,adam_2016, gomes_2017}, and coarser mesh throughout the computational domain, simulation performed with fixed mesh showed larger smeared fluid saturation.

@@

Simulations involving formation and growth of viscous fingers were conducted with the CVFEM-based multi-fluid flow simulator Fluidity/IC-FERST model. The model is based upon a dual consistent pressure-velocity representation in CV and FEM spaces and a novel family of FE-pairs, \PN[n]{m} and \PNDG[n]{m}. Saturation and other scalar fields are discretised in CV space and may be fully discontinuous across finite element boundaries. In order to strongly enforce discretised Darcy equations at the boundaries between elements and control volumes, a directional-weighted flux-limited was introduced to take into account discontinuities risen from both, numerical formulation (based on discontinuous-Galerkin finite element method, DGFEM) and control volumes within finite elements.

%(2)Summary (1-3 paragraphs) of the main results with the appropriate analysis (\ie how you work has proved the theory);
The numerical formulation was initially validated against laboratory experiments \citep{dawe_2008} to qualitatively assess model functionality to simulate crossflow (\ie preferential flow pathways) in a chequerboard permeability domain. Formation and development of viscous fingers in Hele-Shaw cells (homogeneous domain) were investigated to demonstrate that the onset of interface flow instability occur when VR$\ge$3. Maximum dendritic finger widths obtained from numerical simulations performed with VR=10 and VR=150 largely agree with expected parametric analytic solutions \citep{mclean_1981,guan_2003}. However, when perturbations are imposed into the domain's geophysical properties, \ie spatial variability of absolute permeability, numerical solutions partially agree with analytic solutions (Table~\ref{table:Heleshaw}). This may be due to the domain homogeneity assumption in the analytic solution (Eqns.~\ref {eqn:guaneqn}).

Determining the early-onset flow instability is critical to accurately simulate formation, growth and coalescence of fingers, and therefore to: (a) predict displacement (sweep) efficiency and (b) design strategies to either mitigate or avoid fingering and preferential flow pathways. Mesh resolution is thus crucial to investigate Saffmann-Taylor instabilities, and adaptive dynamic mesh technology has proved to be an efficient tool to capture initial interface perturbations. Numerical simulations were performed with VR=10 using fixed and adaptive \PN[1]{2} element-pairs. Results (Figs.~\ref{fig:5regions_plottimenodeselements} and~\ref{fig:5regions_AdaptZoom}) demonstrated that by dynamically adapting the mesh to follow saturation gradients (\ie fluids interfaces), early-onset flow instability and further formation of long multi-fluid fingers, which have themselves split into sub-fingers, can be readily captured (see Figs.~\ref{fig:5regions_fixedmesh}i and~\ref{fig:5regions_adaptmesh}i). Overall computational cost of the simulation conducted with dynamically adapting the mesh were significantly smaller than the simulation conducted with fixed mesh as indicated by the number of elements/nodes (Figs.~\ref{fig:5regions_plottimenodeselements}).

Investigation of preferential flow pathways was the focus of Section~\ref{section:results_3D}, where 3D numerical simulations of multi-fluid flow through semi-pervious / impervious channelised domain were performed. Numerical data indicated preferential flow through regions of larger permeability leaving adjacent region (with low permeability) with nearly stagnant fluids (Figs.~\ref{fig:3DChannel_sat}-~\ref{fig:3DChannel_planes}). Order of accuracy of the element-pair and mesh resolution play key-roles to capture directional flow pathways and fingers in 3D simulations.
\medskip

% (3)(optional) 1 paragraph of potential future work. Make sure that (1) and (2) are covered in the present Section.
This work focuses on complexity associated with flow channelling and instability dynamics in heterogeneous porous media. Methods used to simulate such flows are based in a novel high-order CVFEM formulation that accurately preserves sharp fluid saturation gradients associated with contrasting permeability distribution. Future work will include numerical investigation of (a) crossover from capillary fingering (at low fluid velocity) to viscous fingering for immiscible unstable flows, and (b) coupled viscous and density fingering in miscible fluid flows.

@@

\clearpage 
%% References with bibTeX database:
\bibliographystyle{elsarticle-harv} 
\bibliography{references_Feb}

@@

\listoftables
\clearpage
\input{article_table_Feb}
\clearpage  
\listoffigures
\clearpage

\input{article_figure1_Feb}
%\input{article_figure2}
